Reduced plasticity and variance in physiological rates for ectotherm populations under climate change

### 2024-02-09 ms.qmd
Fixed a bunch of spelling mistakes in new section.

Understanding the interplay between plasticity and genetic adaptation has important implications for predicting a populations resilience to climate change [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate; @bush2016incorporating; @seebacher2023can; @cooke2021one]. For example, incorperating phenotypic plasticity and adaptation into species distribution models has been shown to dampen the predicted contraction of distributions in the face of climate change [e.g., @bush2016incorporating]. In addition, it may be expected that reductions in variance impact the 'opportunity for selection' by reducing the strength of selection and/or the genetic variance exposed to selection, impacting a population's capacity to evolve under climate change [@hoffmann2011climate; @urban2023and]. Quntifying the degree of plasticity and genetic/phenotypic variation in key physiolgical responses is therefore recognised as being critical to informing projections for organisms threatened by climate change [@cooke2021one].

@@

Our findings highlight the potential vulnerability of terrestrial and marine ectotherms to climate change. Our meta-analytic results provide opportunities to parameterise predictive models in ways that capture the percentage changes in relative variances and acclimation responses when predicting species distributions or modelling population resilience in the face of climate warming. Our results also define the range of biological responses observed across diverse taxa (e.g., through prediction intervals) providing opportunities to incorperate realistic biological variation into the modelling process. Incorperating relavant variation in models is needed fully appreciate the biological consequences such changes in variation and plasticity have on ecological and evolutionary dynamics [@bush2016incorporating; @demarche2019incorporating] and are yet to be explored.

@@

Theoretical evolutionary models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly if oscillations are predictable over time, making environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial ecosystems [@steele2019comparison] suggests that plasticity should be more likely to evolve in terrestrial environments. However, we did not find evidence to this effect. This finding is consistent with other meta-analyses [@barley2021limited; @Gunderson2015-hz] with thermal tolerance.  The lack of relationship between physiological plasticity and environmental variability and predictability could be due to the challenges in estimating realistic microthermal environments given that air temperatures may not represent thermal enviroments many taxa actually experience in nature. Therefore, it is possible that more fine-scale resolution of the habitats that organisms occupy will elucidate expected patterns. Randomly fluctuating environments have been suggested to select for reduced plasticity, particularly if plastic responses are costly and this has been demonstrated by a number of empirical studies [@Leung2020; @rescan2022experimental; @leung2023phenotypic]. For example, using seed beetles (*Callosobruchus maculatus*), Hallsson *et al.* @hallsson2012selection showed that experimentally evolving populations under randomly fluctuating thermal conditions did not show any increase in plasticity, but had reduced plasticity. Leung *et al.* @Leung2020 also showed reduced plasticity in morphological traits of experimentally evolving algae populations (*Dunaliella salina*) when environments were less predictable [see also @leung2023phenotypic]. These findings, in combination with our results, suggest that there are costs to being plastic or that the environmental signals are insufficient to trigger endocrine and epigenetic mechanisms that lead to plasticity when environments are not predictable [@Leung2020; @hendry2016key]. The fact that aquatic ectotherms did show a capacity to acclimate suggests that environmental predictability, which is expected to be higher in aquatic environments, may be more important than environmental variability in driving the evolution of plasticity.

>>>>

Understanding the interplay between plasticity and genetic adaptation has important implications for predicting a populations resilience to climate change [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate; @bush2016incorporating; @seebacher2023can; @cooke2021one]. For example, incorporating phenotypic plasticity and adaptation into species distribution models has been shown to dampen the predicted contraction of distributions in the face of climate change [e.g., @bush2016incorporating]. In addition, it may be expected that reductions in variance impact the 'opportunity for selection' by reducing the strength of selection and/or the genetic variance exposed to selection, impacting a population's capacity to evolve under climate change [@hoffmann2011climate; @urban2023and]. Quantifying the degree of plasticity and genetic/phenotypic variation in key physiological responses is therefore recognised as being critical to informing projections for organisms threatened by climate change [@cooke2021one].

@@

Our findings highlight the potential vulnerability of terrestrial and marine ectotherms to climate change. Our meta-analytic results provide opportunities to parameterise predictive models in ways that capture the percentage changes in relative variances and acclimation responses when predicting species distributions or modelling population resilience in the face of climate warming. Our results also define the range of biological responses observed across diverse taxa (e.g., through prediction intervals) providing opportunities to incorporate realistic biological variation into the modelling process. Incorporating relevant variation in models is needed fully appreciate the biological consequences such changes in variation and plasticity have on ecological and evolutionary dynamics [@bush2016incorporating; @demarche2019incorporating] and are yet to be explored.

@@

Theoretical evolutionary models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly if oscillations are predictable over time, making environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial ecosystems [@steele2019comparison] suggests that plasticity should be more likely to evolve in terrestrial environments. However, we did not find evidence to this effect. This finding is consistent with other meta-analyses [@barley2021limited; @Gunderson2015-hz] with thermal tolerance.  The lack of relationship between physiological plasticity and environmental variability and predictability could be due to the challenges in estimating realistic microthermal environments given that air temperatures may not represent thermal environments many taxa actually experience in nature. Therefore, it is possible that more fine-scale resolution of the habitats that organisms occupy will elucidate expected patterns. Randomly fluctuating environments have been suggested to select for reduced plasticity, particularly if plastic responses are costly and this has been demonstrated by a number of empirical studies [@Leung2020; @rescan2022experimental; @leung2023phenotypic]. For example, using seed beetles (*Callosobruchus maculatus*), Hallsson *et al.* @hallsson2012selection showed that experimentally evolving populations under randomly fluctuating thermal conditions did not show any increase in plasticity, but had reduced plasticity. Leung *et al.* @Leung2020 also showed reduced plasticity in morphological traits of experimentally evolving algae populations (*Dunaliella salina*) when environments were less predictable [see also @leung2023phenotypic]. These findings, in combination with our results, suggest that there are costs to being plastic or that the environmental signals are insufficient to trigger endocrine and epigenetic mechanisms that lead to plasticity when environments are not predictable [@Leung2020; @hendry2016key]. The fact that aquatic ectotherms did show a capacity to acclimate suggests that environmental predictability, which is expected to be higher in aquatic environments, may be more important than environmental variability in driving the evolution of plasticity.
